Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-11 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
Section V.B

We now consider dCS gravity.
%
In Fig.~\ref{fig:dCS_exec_sum} we show the marginalized posterior distribution
function of $\ell_{\rm CS}$ for both events GW150914 (top panel) and GW200129
(middle panel), with different colors distinguishing cases with and without spin corrections to
dominant $(2,2,0)$~QNM.
%
Perhaps the most salient feature of these posteriors is that they are not
peaked at $\ell_{\rm CS}$, which may mislead us to believe that we are seeing a
deviation from GR. This is not the case and the situation is that same as the
one we discussed in the previous section for EdGB gravity.
%
We verified that
%
\begin{equation}
    \gamma_{\rm CS} = \left( \frac{c^2 \ell_{\rm CS}}{G M_{f}} \right)^{4}\,,
\end{equation}
%
does indeed peak at zero indicating consistency with GR, similarly to what is shown in Fig.~\ref{fig:gamma_gb} for $\gamma_{\rm GB}$.
%
We also see that in both cases that the inclusion leading-order-in-spin
correction to the mode displaces the posteriors towards smaller values of
$\ell_{\rm CS}$. This can be seen more evidently by looking at the location of
posterior peaks.
%
Finally, in the bottom panel, we show the combined result for both events.

In Fig.~\ref{fig:dcs_cdf} we show the CDF
obtained from Eq.~\eqref{eq:def_cdf}, with $\ell_{\rm max} = \Lambda_{\rm EFT} (\varepsilon, \gm)$,
for both GW150914 (top panel) and GW200129 (bottom panel), where $\Lambda_{\rm EFT}$
is calculated with either $\gm = m_2$ (dashed lines) or $M_{f}$ (solid lines),
and in the range $\varepsilon \in [0,1]$.
%
For GW150914, we see that with $\gm = m_2$ that Eq.~\eqref{eq:eft_bound} is not
satisfied unless $\varepsilon \approx 0.9$ (with only $n=0$ corrections) and
$\varepsilon \approx 0.7$ (with both $n=0$ and $1$ corrections).
%
The situation is different if we use $\gm = M_f$. In this case, we find that
with or without spin corrections that Eq.~\eqref{eq:eft_bound} can be satisfied
with $\varepsilon \leqslant 1/2$, i.e., below the criteria used
Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
This means that with our model's assumptions and using the remnant's source mass $M_f$ to set the
cut-off scale that we can claim an upper bound
%
\begin{equation}
\ell_{\rm CS} \leqslant 42.6~\textrm{km}
\quad \textrm{(90\% credibility)},
\end{equation}
%
on dCS gravity and would constitute the first bound on this theory with
gravitational-wave observations alone.
%
We also note that since the CDF grows fast in approximate domain $\varepsilon = [0.2, 0.4]$,
that a stronger, albeit at lower credibility,
%
\begin{equation}
\ell_{\rm CS} \leqslant 37.4~\textrm{km}
% ~( = \Lambda_{\rm EFT}(0.3, M_{f}) )
\quad \textrm{at 68\% credibility},
\end{equation}
%
can also be placed, when $N_{\rm max}=0$.

We can draw qualitatively similar conclusions from the GW200129 event.
In particular, we find,
%
\begin{equation}
\ell_{\rm CS} \leqslant 36.6~\textrm{km}
% 43.9~\textrm{km}~( = \Lambda_{\rm EFT}(\tfrac{1}{2}, M_{f}) )
\quad \textrm{(90\% credibility)},
\end{equation}
%
and
%
\begin{equation}
\ell_{\rm CS} \leqslant 29.1~\textrm{km}
% ~( = \Lambda_{\rm EFT}(0.3, M_{f}) )
\quad \textrm{(68\% credibility)},
\end{equation}
%
when $N_{\rm max}=0$.
%
These stronger bounds are a consequence of the larger support for $\ell_{\rm CS} \lessapprox 15$~km
for GW200129 (compare the top and middle panels in Fig.~\ref{fig:dCS_exec_sum}), and in part due to
the smaller median remnant ($M_{f} \approx 59.5\msun$ versus $M_{f} \approx 61.8\msun$ for GW150914).
%
\hs{Should we remove these and just quote the 90\% results?}

@@

Finally, since both event individually led to a bound on $\ell_{\rm CS}$
(assuming a cut-off scale for $\gm = M_{f}$ and $j = 0$), we can combine the
posteriors to obtain the cumulative bound,
%
\begin{equation}
\ell_{\rm CS} \leqslant 38.7~\textrm{km}
\quad \textrm{(90\% credibility)},
\end{equation}
%
which is the main result of this section.
%
This bound is approximately a factor of four weaker than that placed by
Ref.~\cite{Silva:2020acr}, but
%
(i) relies only on GW observations and
%
(ii) suggests that a ringdown analysis can potentially place constraints on
theories which evade GW tests using inspiral information alone, such as the
case of dCS gravity~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
In Table~\ref{tab:summary_dcs} we summarize our findings of this section.

\begin{table}[h]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event & EFT    & ParSpec & Constraint ($\gm = M_{f}$) \\
              &       & bound? & bound?  &                            \\
\hline
%       & GW150914 & Yes & Yes & $\ell_{\rm CS} \leqslant 46.5$~km \\
% 0     & GW200129 & Yes & Yes & $\ell_{\rm CS} \leqslant 43.9$~km \\
%       & Combined & Yes & Yes & \cellcolor{blue!10}$\ell_{\rm CS} \leqslant 34.5$~km \\
      & GW150914 & Yes & Yes & $\ell_{\rm CS} \leqslant 41.9$~km \\
0     & GW200129 & Yes & Yes & $\ell_{\rm CS} \leqslant 35.8$~km \\
      & Combined &     &     & \cellcolor{black!10}$\ell_{\rm CS} \leqslant 38.7$~km \\
\hline
      & GW150914 & Yes & No  & --                                 \\
1     & GW200129 & Yes & No  & --                                 \\
      & Combined &     &     & --                                 \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results for dCS gravity for GW150914, GW200129, and
combined events using $\gm = M_{f}$, $\varepsilon = 1/2$ and quoting only 90\% credible results.
%
We found that while our posteriors satisfy the condition~\eqref{eq:eft_bound} (with $\varepsilon = 1/2$),
they do not obey the condition~\eqref{eq:parspec_bound} for $j = 1$. This means
that our results for $N_{\rm max}=0$ are the only ones we can confidently quote.
%
The combined bound, which is also quoted in Table~\ref{tab:ref_theories_qnms},
is $\ell_{\rm CS} \leqslant 38.7$~km.
}
\label{tab:summary_dcs}
\end{table}

>>>>

We now consider dCS gravity \ag{where the main results are summarised in Fig.~\ref{fig:dCS_exec_sum}}.
%
\sout{In Fig.~\ref{fig:dCS_exec_sum} we show the marginalized posterior distribution
function of $\ell_{\rm CS}$ for both events GW150914 (top panel) and GW200129
(middle panel), with different colors distinguishing cases with and without spin corrections to
dominant $(2,2,0)$~QNM.}\agcomm{A lot of the language here is a repetition of the previous section.}
%
\sout{Perhaps the most salient feature of these posteriors is that they are not
peaked at $\ell_{\rm CS}$, which may mislead us to believe that we are seeing a
deviation from GR. This is not the case and the situation is that same as the
one we discussed in the previous section for EdGB gravity.}
\ag{As with EdGB gravity (Sec.~\ref{sec:results_edgb}) although the PDF of $\ell_{\rm CS}$
is peaked away from 0, this does not signify a deviation from GR, as we have verified that,}
%
\sout{We verified that}
%
\begin{equation}
    \gamma_{\rm CS} = \left( \frac{c^2 \ell_{\rm CS}}{G M_{f}} \right)^{4}\,,
\end{equation}
%
does indeed peak at zero indicating consistency with GR, similarly to what is shown in Fig.~\ref{fig:gamma_gb} for $\gamma_{\rm GB}$.
%
We also see that in both cases that the inclusion leading-order-in-spin
correction to the mode displaces the posteriors towards smaller values of
$\ell_{\rm CS}$. This can be seen more evidently by looking at the location of
posterior peaks.
%
Finally, in the bottom panel, we show the combined result for both events.
\agcomm{Just a thought.  Since, Figs 3,6,8,10 and Figs.4,6,9,11 show the same basic thing,
should we have one description at the top of the section for each figure? We can follow this up
by what is special for each theory in each section? Currently, we are repeating a lot of the 
language to describe these figures for each theory?}

In Fig.~\ref{fig:dcs_cdf} we show the CDF
\sout{obtained from Eq.~\eqref{eq:def_cdf}, with $\ell_{\rm max} = \Lambda_{\rm EFT} (\varepsilon, \gm)$,
for both GW150914 (top panel) and GW200129 (bottom panel), where $\Lambda_{\rm EFT}$
is calculated with either $\gm = m_2$ (dashed lines) or $M_{f}$ (solid lines),
and in the range $\varepsilon \in [0,1]$.}
%
\sout{For}\ag{for} GW150914, we see that with $\gm = m_2$ that Eq.~\eqref{eq:eft_bound} is not
satisfied unless $\varepsilon \approx 0.9$ (with only $n=0$ corrections) and
$\varepsilon \approx 0.7$ (with both $n=0$ and $1$ corrections).
%
The situation is different if we use $\gm = M_f$. In this case, we find that
with or without spin corrections that Eq.~\eqref{eq:eft_bound} can be satisfied
with $\varepsilon \leqslant 1/2$, i.e., below the criteria used
Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
This means that with our model's assumptions and using the remnant's source mass $M_f$ to set the
cut-off scale that we can claim an upper bound
%
\begin{equation}
\ell_{\rm CS} \leqslant 42.6~\textrm{km}
\quad \textrm{(90\% credibility)},
\end{equation}
\agcomm{I am not familiar with the term ``credibility", and am more used to the phrase, ``at X\% credible level". 
Are they the same thing?}
%
on dCS gravity and would constitute \emph{the first bound on this theory with
gravitational-wave observations alone}.\agcomm{Since this is an extremely important
result, would it make sense to reorder the theories such that dCS comes first among results?}
%
We also note that since the CDF grows fast in approximate domain $\varepsilon = [0.2, 0.4]$,
that a stronger, albeit at lower credibility,
%
\begin{equation}
\ell_{\rm CS} \leqslant 37.4~\textrm{km}
% ~( = \Lambda_{\rm EFT}(0.3, M_{f}) )
\quad \textrm{at 68\% credibility},
\end{equation}
%
can also be placed, when $N_{\rm max}=0$.

We can draw qualitatively similar conclusions from the GW200129 event.
In particular, we find,
%
\begin{equation}
\ell_{\rm CS} \leqslant 36.6~\textrm{km}
% 43.9~\textrm{km}~( = \Lambda_{\rm EFT}(\tfrac{1}{2}, M_{f}) )
\quad \textrm{(90\% credibility)},
\end{equation}
%
and
%
\begin{equation}
\ell_{\rm CS} \leqslant 29.1~\textrm{km}
% ~( = \Lambda_{\rm EFT}(0.3, M_{f}) )
\quad \textrm{(68\% credibility)},
\end{equation}
%
when $N_{\rm max}=0$.
%
These stronger bounds are a consequence of the larger support for $\ell_{\rm CS} \lessapprox 15$~km
for GW200129 (compare the top and middle panels in Fig.~\ref{fig:dCS_exec_sum}), and in part due to
the smaller median remnant ($M_{f} \approx 59.5\msun$ versus $M_{f} \approx 61.8\msun$ for GW150914).
%
\hs{Should we remove these and just quote the 90\% results?}\agcomm{I think
it's fine to give point estimates, because it's easier for reader to interpret results 
rather than having to make sense of error bars.}

@@

Finally, since both events individually led to a bound on $\ell_{\rm CS}$
(assuming a cut-off scale for $\gm = M_{f}$ and $N_{\rm max}=0$), we can combine the
posteriors to obtain the cumulative bound,
%
\begin{equation}
\ell_{\rm CS} \leqslant 38.7~\textrm{km}
\quad \textrm{(90\% credibility)},
\end{equation}
%
which is the main result of this section.
%
This bound is approximately a factor of four weaker than that placed by
Ref.~\cite{Silva:2020acr}, but
%
(i) relies only on GW observations and
%
(ii) suggests that a ringdown analysis can potentially place constraints on
theories which evade GW tests using inspiral information alone, such as the
case of dCS gravity~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
In Table~\ref{tab:summary_dcs} we summarize our findings of this section.

\begin{table}[h]
\begin{tabular}{l l c c c}
\hline
\hline
$N_{\rm max}$ & Event & EFT    & ParSpec & Constraint ($\gm = M_{f}$) \\
              &       & bound? & bound?  &                            \\
\hline
%       & GW150914 & Yes & Yes & $\ell_{\rm CS} \leqslant 46.5$~km \\
% 0     & GW200129 & Yes & Yes & $\ell_{\rm CS} \leqslant 43.9$~km \\
%       & Combined & Yes & Yes & \cellcolor{blue!10}$\ell_{\rm CS} \leqslant 34.5$~km \\
      & GW150914 & Yes & Yes & $\ell_{\rm CS} \leqslant 41.9$~km \\
0     & GW200129 & Yes & Yes & $\ell_{\rm CS} \leqslant 35.8$~km \\
      & Combined &     &     & \cellcolor{black!10}$\ell_{\rm CS} \leqslant 38.7$~km \\
\hline
      & GW150914 & Yes & No  & --                                 \\
1     & GW200129 & Yes & No  & --                                 \\
      & Combined &     &     & --                                 \\
\hline
\hline
\end{tabular}
\caption{Detailed summary of our results for dCS gravity for GW150914, GW200129, and
combined events using $\gm = M_{f}$, $\varepsilon = 1/2$ and quoting only 90\% credible bounds.
%
We found that while our posteriors satisfy the condition~\eqref{eq:eft_bound} (with $\varepsilon = 1/2$),
they do not obey the condition~\eqref{eq:parspec_bound} for $N_{\rm max} = 1$. This means
that our results for $N_{\rm max}=0$ are the only ones we can confidently quote.
%
The combined bound, which is also quoted in Table~\ref{tab:ref_theories_qnms},
is $\ell_{\rm CS} \leqslant 38.7$~km.
}
\label{tab:summary_dcs}
\end{table}
